\color{blue} On the Synchronization of Computational RFIDs

### 2017-10-25 manuscript_tmc.tex
TMC submission version

\documentclass[journal,draftcls,onecolumn,12pt,twoside]{IEEEtranTCOM}

\normalsize
%\documentclass[english,journal]{IEEEtran}

% Metadata Information
%\acmVolume{9}
%\acmNumber{4}
%\acmArticle{39}
%\acmYear{2016}
%\acmMonth{3}

% Copyright
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

% DOI
%\doi{0000001.0000001}

%ISSN
%\issn{1234-56789}

%\usepackage[latin9]{inputenc}
%\usepackage{hyphenat}
%\usepackage{epstopdf}
%\usepackage{flushend}
\usepackage{algorithm,setspace}
\usepackage[noend]{algpseudocode} % needed for: algorithm pseudocode
\usepackage{color}
\usepackage{array}
\usepackage{multirow}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{url}
\usepackage{hyperref}
\usepackage{stmaryrd}
\usepackage{soul}
\usepackage{cancel}
\usepackage{listings}
\usepackage{tikz} % needed for: tikz pics
\usetikzlibrary{arrows}
\usetikzlibrary{decorations.pathreplacing}

@@

\newcommand{\expected}[1]{E\left\llbracket #1 \right\rrbracket}
\newcommand{\revised}[1]{\textcolor{blue}{#1}}
\newcommand{\TODO}[1]{\textcolor{blue}{\textbf{TODO: }}\textcolor{red}{#1}}
\newtheorem{fact}{Fact}

% text revision commands
%\newcommand{\add}[1]{\textcolor{blue}{#1}}
%\newcommand{\remove}[1]{\textcolor{red}{\st{#1}}}
%\newcommand{\replace}[2]{\remove{#1} \add{#2}}

%\makeatother

% Document starts
\begin{document}

% Title portion
\title{On the Synchronization of Intermittently Powered Wireless Embedded Systems}

\author{Kas{\i}m~Sinan~Y{\i}ld{\i}r{\i}m, Henko~Aantjes, Przemys{\l}a~Pawe{\l}czak and Amjad~Yousef~Majid
%
\thanks{A preliminary version of this article was presented at the HLPC 2016 Workshop, Atlanta, GA, USA, April 2, 2016~\cite{hlpc2016}.}%
\thanks{The authors are with the Embedded Software Group, Delft University of Technology, Mekelweg 4, 2628 CD Delft, The Netherlands (e-mail: \{k.s.yildirim, p.pawelczak, a.y.majid\}@tudelft.nl, h.aantjes@student.tudelft.nl).}}

\markboth{IEEE Transactions on Communications}%
{Submitted paper}

@@

\begin{abstract}
Battery-free computational RFID platforms, such as WISP (Wireless Identification and Sensing Platform), are intermittently-powered devices designed for replacing existing sensor networks. Accordingly, synchronization appears as one of the  crucial building blocks for collaborative and coordinated actions in these platforms. However, intermittent power leads to frequent loss of computational state and short-term clock frequency instability that makes synchronization challenging. In this article, we introduce WISP-Sync protocol that provides synchronization among WISP tags in the communication range of an RFID reader. WISP-Sync overcomes the aforementioned challenges by employing a Proportional-Integral (PI) controller-inspired algorithm which (i) is adaptive---reactive to short-term clock instabilities; (ii) requires only a few computation steps---suitable for limited harvested energy; and (iii) keeps a few variables to hold the synchronization state---minimum overhead to recover from power interrupts.	Evaluations in our testbed showed that WISP-Sync ensured an average synchronization error of approximately 1\,ms among the tags with an average energy overhead of 1.85\,$\mu$J per synchronization round.
\end{abstract}
%
\begin{IEEEkeywords}Time Synchronization, Computational RFIDs, Wireless Identification and Sensing Platform (WISP).
\end{IEEEkeywords}

@@

\section{Introduction}
\label{sec:Intro}

\bstctlcite{BSTcontrol}\IEEEPARstart{L}{ow-power} wireless embedded systems consist of tiny, low-cost and spatially separated computers that communicate with each other by exchanging radio packets. Powering these small-scale embedded systems, e.g. wireless sensor networks (WSNs), is still a crucial problem~\cite{emergence}. Replacing or recharging their batteries is impractical and inhibits long-term operation. Moreover, batteries increase the size and cost of the hardware. Fortunately, the energy efficiency of these systems has improved considerably such that their power requirements are in the order of a few $\mu$W \cite{booksmith2013}. Furthermore, recent advancements in microelectronics technology enabled harvesting power from radio frequency (RF) sources that is sufficient to operate low-power embedded systems in practice~\cite{60ghz,emergence,bi:commag:2015,xie_wcom_2013,garnica_procieee_2013}. Nowadays, the growth of the RF-powered computing paradigm is bringing new research opportunities and challenges~\cite{emergence}, leading to a promising class of low-power embedded systems, the so-called Intermittently Powered Devices (IPDs).

@@

 As of now, battery-less networks are in the form of one-hop architecture in which the CRFID tags communicate only with the RFID reader~\cite{wisp,booksmith2013,wisent:2016}. However, recent advancements in the wireless communication enable tag-to-tag communication, e.g.~\cite{liu2013ambient}. We anticipate that these multi-hop networks will demand their own implementation of basic sensor network building blocks~\cite{levis:tinyos} with the aim of replacing the existing battery-powered sensing and data collection applications with their battery-less counterparts. In turn, we conjecture, that in this new battery-less sensing architecture the synchronization will be a fundamental building block that allows, for instance, temporal ordering of the collected data in order to make inferences.

%\add{\textbf{Distributed Sensing.} As an example, consider applications of monitoring inaccessible areas or general surveillance in which multiple battery-less cameras, such as WISPCam~\cite{wispcam_2015}, capture images from different locations simultaneously and forward the collected data to the RFID reader. In such applications, the WISPCams can coordinate their image capture actions by means of the almost simultaneously received periodic ``\texttt{perfrom image snapshot}'' commands from the RFID reader. However, the reception and processing of the commands, and in turn the coordination, can be interrupted since WISPCams might exhibit different \emph{charge/discharge} patterns. In principle, WISPCams operate in a power cycling mode, in which their storage capacitor is fully-charged in the charging phase and then discharged until a lower threshold (no operation) voltage is attained. Due to their different distances to the RFID reader leading to different harvested energy levels, it can happen that a WISPCam would have to wait until it is fully-charged to capture the image. It has been observed that ``the charging time for voltage to reach 1.8\,V to power a WISP equipped with a 100\,$\mu$F capacitor can be as large as 155 seconds, when the RFID reader is 10.0 meters away''~\cite{fu2013minimizing}. This extra delay might result in the loss of coordination and the capture of images at different time instants. To prevent this situation, the reader can send ``\texttt{perform image snapshot $x$ seconds later}' command and continue charging the WISPCams. Then, WISPCams can activate themselves and capture the image simultaneously, just after $x$ seconds after the reception of the reader command. Since each WISPCam has its own clock hardware whose oscillator generate pulses at slightly different speeds, a \emph{synchronization} service that allows WISPCams to wake-up simultaneously and facilitates such collaborative and coordinated actions becomes a fundamental requirement for these systems.}

%that the charging time for voltage to reach 1.8V to power a WISP equipped with a 100uF capacitor can be as large as 155 seconds, when the RFID reader is 10.0 meters away

%It should be noted that there can also be a degree of synchronization on the 
%reader side by timestamping the received 
%information. However, the WISPCams that experience a power 
%loss might resume transmission of the captured image to the reader after a 
%relatively long delay due to harvesting sufficient energy. In this case, the 
%timestamps do not make sense since they do not reflect the actual capture time 
%of the images. Therefore, the establishment of a ``common time'' notion

\label{sec:Motivation}

The characteristics of CRFID systems expose fundamentally different challenges than sensor networks to implement the synchronization service. The reason is mainly twofold:

\begin{itemize}
\item \textbf{Challenge I:} CRFID systems should perform computations in an energy efficient manner despite intermittent RF power that leads to frequent loss of computational state, e.g. when the RFID reader moves away from the CRFID tag \cite{ransford:2008}.

\item \textbf{Challenge II:} The continuously varying voltage supply introduces severe hardware instability, e.g. varying oscillator frequencies in short-term affects the stability of the clocks and degrades the accuracy sensing.	
\end{itemize}

The focus of this article is to answer the question of \emph{how to design a building block that synchronizes intermittently-powered wireless embedded systems?} To this end, we investigate the WISP platform and provide initial observations and limitations pertaining to the synchronization of these devices. In particular, the main contributions and findings of 
this article are:
%
\begin{itemize}
\item To the best of our knowledge, this is the first study that focuses on the synchronization of tags in CRFID systems. Even though there are studies on the synchronization of multiple RFID readers, e.g.~\cite{reader_sync}, we are unaware of any study that provides explicit synchronization of individual tags in the communication range of an RFID reader. 
\item By addressing the aforementioned challenges, we design and implement\emph{the first synchronization primitive} for the tags, namely \emph{WISP-Sync}, inspired by the Proportional-Integral (PI) controllers~\cite{pi2015}. WISP-Sync (i) 
requires only a few computation steps to run efficiently under limited harvested energy and keeps a few variables to hold the synchronization state to recover from power interruptions with minimum overhead---\emph{addressing Challenge I}; (ii) is adaptive to react to clock instabilities in a fast manner---\emph{addressing Challenge II}.
%
\item We provide theoretical analysis to prove that this primitive establishes synchronization and to reveal the factors affecting its synchronization performance.
%
\item Evaluations in our testbed show that an average synchronization error of approximately 1\,ms can be ensured among tags with an energy overhead of 1.85\,$\mu$J per synchronization round using this  primitive.
\end{itemize}
%
We note that WISP-side implementations throughout the article are done using C and assembly language. All source codes and scripts used to generate the results in the article (including parsing, post-processing and measurement results) are available via~\href{https://goo.gl/RuWIZ2}{https://goo.gl/RuWIZ2}.

@@

\section{Synchronizing Battery-Free Wireless Embedded Systems}
\label{sec:Related-Work}

@@

\subsection{Synchronization in Conventional Wireless Sensor Networks}
\label{sec:WSN-Sync}

@@

\subsubsection{Computation Methods}

The most common synchronization mechanism is to propagate the time information of a particular \emph{reference node} to let receiver nodes synchronize themselves to the received reference time information. Commonly, \emph{least-squares regression} is employed by several practical synchronization protocols to adjust the offset and the frequency of the software clocks ~\cite{Maroti2004,Lenzen:2015,TATS:2016}. On the other hand \emph{iterative} computation methods, e.g. in ~\cite{Yildirim:AVTS:2014,pi2015} and \cite{Yildirim:Gradient:2016}, perform identical computation steps at each re-synchronization round. With iterative methods, sensor nodes establish synchronization after a finite number of rounds in a lightweight manner in terms of computation and memory requirements. There are also \emph{fully distributed} approaches in which sensor nodes interact only with and synchronize to their direct neighbors in a peer-to-peer fashion without the requirement of a dedicated reference node. In these approaches, nodes employ computation methods based on \emph{distributed consensus}~\cite{Sommer2009Gradient,SchenatoFiorentin:2011,Yildirim:External:2014}.

\subsection{Fundamental Challenges of Synchronizing IPDs: The CRFID Case}
\label{sec:CRFIDs}

We now delve into the synchronization characteristics of RF-powered wireless embedded systems. To this end, we will consider the WISP, the de facto CRFID platform. The main aspects pertaining to synchronization in the WISP platform can be stated as follows:

\paragraph{Single-hop reader-tag architecture} WISP tags are deployed inside the communication range of an RFID reader and they can communicate only with the reader using backscatter communication. Therefore, the RFID reader itself is the natural reference device to establish synchronization among the WISPs, promoting \emph{reader-tag synchronization}. Since WISP tags are unable to communicate with their neighboring nodes directly using backscatter communication as of now, \emph{tag-tag synchronization} is impossible.

\begin{figure}
	\centering
	\includegraphics[scale=0.35]{figures/clock_frequency_dc_rf.eps}
	\caption{\label{fig:frequency_dc_rf} In our testbed, see Section~\ref{sec:testbed} for the details, we programmed a single WISP tag so that it toggles one of its output port whenever its 32~KHz built-in timer progresses 32000 clock ticks. In the ideal case where the crystal oscillator of the tag does not drift, the tag would toggle its output port at each 1 second, i.e. with a frequency of 1\,Hz. During our experiment under a stable environmental temperature, we connected the output port of the tag to a logic analyzer and measured the frequency of the toggle event. We plotted our measurements when the tag is powered by the Flash Emulation Tool (FET) with a stable DC power and when the tag is powered by the RFID Reader at a distance of 5~cm. Since harvested RF energy was different and unstable than the constant DC case, the frequency of the oscillator and in turn the toggle event was quite different. We measured the mean and standard deviation of the frequency in the DC case as 1.1653 and 0.00028553, and these values for the RF case were 1.1793 and 0.00026049, respectively. }
\end{figure}

\paragraph{Continuously varying voltage level} In WSNs, the battery level decreases gradually that allows stable voltage levels in the short term. On the contrary, on the WISP platform, the fluctuating input voltage prevents short-term stability of the clock hardware and introduces significant drift. Hence, the prominent factor affecting the frequency of the crystal oscillator is the varying voltage level rather than the temperature of the WISP tags, see Fig.~\ref{fig:frequency_dc_rf}.

\paragraph{Frequent loss of synchronization state} Contrary to sensor nodes, WISP tags frequently ``die'' due to power loss and they need to save the synchronization state, e.g. the clock offset and relative frequency, into the non-volatile memory to recover when they harvested sufficient energy to start up. However, saving computational state is also an energy consuming task~\cite{blisp}.

\paragraph{Computation and memory overhead sensitivity} The classical motto of WSNs, ``compute instead of communicate whenever possible''~\cite[p. 44]{karl2007protocols}, is no longer valid for the WISP platform since backscatter communication comes almost for free~\cite{ekhonet}. Due to the intermittent power, lightweight methods in terms of computation and memory are required for the synchronization. Since methods like least-squares regression are computationally heavy and require considerable amount of memory~\cite{pi2015} they should be avoided.

\paragraph{Limitations of the EPC Gen 2 standard} The WISP firmware implements the EPC Gen 2 standard~\cite{epc_gen2} which increases the compatibility with the existing RFID systems. However, the standard introduces limitations, e.g. currently it does not assign timestamps to the radio packets, which is a fundamental requirement to establish synchronization. Moreover, communication delays between the reader and tag are quite dependent on the implementation of this standard by RFID readers. Unfortunately, these issues lead to less accurate synchronization as compared to existing WSN solutions, as shown by our measurements presented in the following sections.

@@

\subsection{Low Power Operation Modes and WISP Clock System}
\label{subsec:clock_system}

@@

\begin{figure}
\centering
	\begin{tikzpicture}
	\node[anchor=south west, inner sep=0] at 
	(0,0){\includegraphics[height=.35\columnwidth]{figures/testbed_resized.jpg}};

	\node [fill= white,circle] at (9,2) { };
	\node [color=black ] at (9,2) { 1 };

	\node [fill= white,circle] at (9,4) {  };
	\node [color=black ] at (9,4) { 2 };

@@

	\node [fill= white,circle] at (5.5,2) {  };
	\node [color=black ] at (5.5,2) { 4 };

	\node [fill= white,circle] at (8,5) {  };
	\node [color=black ] at (8,5) { 5 };

@@

In this section, we provide a mathematical framework that models the built-in clocks and in particular the clock drifts of the WISP tags and the RFID reader. We will use these models to prove the correctness of our synchronization algorithms and to derive mathematical expressions that represent their synchronization performance.
\begin{figure*}
\centering
\includegraphics[scale=0.35] {figures/relative_clock_frequency.eps}
\caption{\label{fig:frequency}The frequencies of the toggle events (obtained using the same steps presented in Fig.\ref{fig:frequency_dc_rf}) of two RF-powered WISP tags, placed at 5 cm distance from the RFID reader. On the left the individual event frequencies ($f_1$ and $f_2$) are presented, and their relative clock frequency ($\frac{f_1}{f_2}-1$) is presented on the right. We measured the mean and standard deviation of $f_1$ as 1.1793 and 0.00026049 and those of $f_2$ as 1.1705 and  0.00068239, respectively.}
\end{figure*}
In this framework, we denote a single WISP tag by $w$ and the RFID reader by $r$. The Timer B0 mentioned in Section~\ref{subsec:clock_system} can be considered as the built-in clock, i.e. \emph{local clock}, of the tag $w$, 
which we denote by $C_{w}$. The value of $C_{w}$ at time $t$ can be modeled as
%
\begin{eqnarray}
C_{w}(t) & \triangleq & C_w(t_0)+ \int_{t_0}^{t}f_{w}(u)du, 
\label{eq:local_clock}
\end{eqnarray}
%
where $t_0$ represents the time at which tag $w$ is powered on and  $f_{w}(u)$ represents the \emph{instantaneous oscillator frequency} of $C_{w}$ at time $u$. Since the frequency of the oscillators are affected by environmental factors such as temperature and voltage level, they are not stable~\cite{Lenzen2009Optimal} and $f_{w}$ is not a constant but it is \emph{time varying}, see Fig.~\ref{fig:frequency}. From \eqref{eq:local_clock}, it can be noticed that $C_w$ is \emph{unitless}. Indeed, it can be seen as a counter of oscillation events that are produced at a frequency of $f_w$.

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{figures/Sender-Receiver.pdf}
\caption{\label{fig:sender_receiver} The message exchange among the host computer, the RFID reader and the tag for sender-receiver synchronization. The host machine, i.e. a PC,  sends the high level commands to the reader via LLRP \emph{AccessSpec} message and receives the results through an \emph{ASReport}. The RFID reader follows the steps defined in the EPC Gen2 standard: (i) performs a \emph{Handshake} to initialize the communication with the active tag; (ii) performs a \emph{Command Initialization} (Cmd Init) by requesting a random number from the tag (Req\_n) and receiving the random number (newRN); (iii)  performs a \emph{Read} command by sending its request and receiving the timer value. The reader assigns the \emph{FirstSeenTimestamp} to the tag at time $t_1$ and the tag timestamps the command reception event at time $t_2$ with its local clock reading.}
\end{figure}

@@

\paragraph{Transmission Delay and Jitter Measurements}

>>>>

\documentclass[10pt,journal,compsoc]{IEEEtran}

\ifCLASSOPTIONcompsoc
  % IEEE Computer Society needs nocompress option
  % requires cite.sty v4.0 or later (November 2003)
  \usepackage[nocompress]{cite}
\else
  % normal IEEE
  \usepackage{cite}
\fi

\usepackage{algorithm,setspace}
\usepackage[noend]{algpseudocode} % needed for: algorithm pseudocode
\usepackage{color}
\usepackage{array}
\usepackage{multirow}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{url}
\usepackage{hyperref}
\usepackage{stmaryrd}
\usepackage{soul}
\usepackage{cancel}
\usepackage{listings}
\usepackage{tikz} % needed for: tikz pics
\usetikzlibrary{arrows}
\usetikzlibrary{decorations.pathreplacing}

@@

\newcommand{\expected}[1]{E\!\left\llbracket #1 \right\rrbracket}
\newcommand{\revised}[1]{\textcolor{blue}{#1}}
\newcommand{\TODO}[1]{\textcolor{blue}{\textbf{TODO: }}\textcolor{red}{#1}}
\newtheorem{fact}{Fact}

% text revision commands
\newcommand{\add}[1]{\textcolor{blue}{#1}}
\newcommand{\remove}[1]{\textcolor{red}{\st{#1}}}
\newcommand{\replace}[2]{\remove{#1} \add{#2}}

\begin{document}

\title{ \color{blue} On the Synchronization of Computational RFIDs}

\author{Kas{\i}m~Sinan~Y{\i}ld{\i}r{\i}m, 
	Henko~Aantjes, 
	Przemys{\l}a~Pawe{\l}czak~\IEEEmembership{Member,~IEEE,} 
	and~Amjad~Yousef~Majid
	%
	\IEEEcompsocitemizethanks{
		\IEEEcompsocthanksitem A preliminary version of this article was presented at the HLPC 2016 Workshop, Atlanta, GA, USA, April 2, 2016~\cite{hlpc2016}. \protect\\
		\IEEEcompsocthanksitem The authors are with the Embedded Software Group, Delft University of Technology, Mekelweg 4, 2628 CD Delft, The Netherlands (e-mail: \{k.s.yildirim, p.pawelczak, a.y.majid\}@tudelft.nl, h.aantjes@student.tudelft.nl).
	}
%	\thanks{Manuscript received April 19, 2005; revised August 26, 2015.}
}

\IEEEtitleabstractindextext{%
\begin{abstract}
Battery-free computational RFID platforms, such as WISP (Wireless Identification and Sensing Platform), are intermittently-powered devices designed for replacing existing sensor networks. Accordingly, synchronization appears as one of the  crucial building blocks for collaborative and coordinated actions in these platforms. However, intermittent power leads to frequent loss of computational state and short-term clock frequency instability that makes synchronization challenging. In this article, we introduce WISP-Sync protocol that provides synchronization among WISP tags in the communication range of an RFID reader. WISP-Sync overcomes the aforementioned challenges by employing a Proportional-Integral (PI) controller-inspired algorithm which (i) is adaptive---reactive to short-term clock instabilities; (ii) requires only a few computation steps---suitable for limited harvested energy; and (iii) keeps a few variables to hold the synchronization state---minimum overhead to recover from power interrupts.	Evaluations in our testbed showed that WISP-Sync ensured an average synchronization error of approximately 1\,ms among the tags with an average energy overhead of 1.85\,$\mu$J per synchronization round.
\end{abstract}

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
Time Synchronization, Computational RFIDs, Wireless Identification and Sensing Platform (WISP).
\end{IEEEkeywords}}

@@

\IEEEdisplaynontitleabstractindextext

@@

\IEEEraisesectionheading{\section{Introduction}\label{sec:Intro}}

\IEEEPARstart{L}{ow-power} wireless embedded systems consist of tiny, low-cost and spatially separated computers that communicate with each other by exchanging radio packets. Powering these small-scale embedded systems, e.g. wireless sensor networks (WSNs), is still a crucial problem~\cite{emergence}. Replacing or recharging their batteries is impractical and inhibits long-term operation. Moreover, batteries increase the size and cost of the hardware. Fortunately, the energy efficiency of these systems has improved considerably such that their power requirements are in the order of a few $\mu$W \cite{booksmith2013}. Furthermore, recent advancements in microelectronics technology enabled harvesting power from radio frequency (RF) sources that is sufficient to operate low-power embedded systems in practice~\cite{60ghz,emergence,bi:commag:2015,xie_wcom_2013,garnica_procieee_2013}. Nowadays, the growth of the RF-powered computing paradigm is bringing new research opportunities and challenges~\cite{emergence}, leading to a promising class of low-power embedded systems, the so-called Intermittently Powered Devices (IPDs).

@@

{\color{blue}{\subsection{The Need for CRFID Network Synchronization}}}

As of now, battery-less networks are in the form of one-hop architecture in which the CRFID tags communicate only with the RFID reader~\cite{wisp,booksmith2013,wisent:2016}. However, recent advancements in the wireless communication enable tag-to-tag communication, e.g.~\cite{liu2013ambient}. We anticipate that these multi-hop networks will demand their own implementation of basic sensor network building blocks~\cite{levis:tinyos} with the aim of replacing the existing battery-powered sensing and data collection applications with their battery-less counterparts. In turn, we conjecture, that in this new battery-less sensing architecture the synchronization will be a fundamental building block that allows, for instance, temporal ordering of the collected data in order to make inferences \add{at the CRFID tag side}. \add{Therefore, establishing a common time notion among the tags, rather than synchronizing RFID readers only, will be a crucial service.}

{\color{blue}{\subsection{Differences From Battery-Powered WSNs }}}
\label{sec:Motivation}

\add{However, }the characteristics of CRFID systems expose fundamentally different challenges than \add{battery-powered} sensor networks to implement the synchronization service. The reason is mainly twofold:

\begin{itemize}
	\item \textbf{Challenge I:} CRFID systems should perform computations \replace{in an energy efficient manner despite}{efficiently using marginal and} intermittent RF power that leads to frequent loss of computational state, e.g. when the RFID reader moves away from the CRFID tag \cite{ransford:2008}. \add{Therefore, computations pertaining to time synchronization have an extremely short time budget, e.g. should be completed with the other computation tasks within 10--100\,ms in WISP~\cite{Ransford:2011:Mementos}. On the contrary, the synchronization approaches in WSNs, e.g. the de facto protocol \cite{Maroti2004}, keep communication infrequent by pushing computation instead, so that the battery power is used efficiently in order to extend the life-time of the network up to several years.}

	\item \textbf{Challenge II:} The continuously varying voltage supply introduces severe hardware instability, e.g. varying oscillator frequencies in short-term affects the stability of the clocks and degrades the accuracy sensing. \add{Therefore, computations pertaining to time synchronization should be reactive against very-frequent voltage effects; e.g. during super-capacitor discharge. On the contrary, in WSNs the input voltage is constant and the temperature is the main factor affecting the clock frequencies~\cite{schmid2010interaction}. Temperature compensation techniques, e.g. \cite{Schmid:2009}, are not applicable to compensate for the voltage instabilities in CRFIDs due to their computational complexity.}

\end{itemize}

{\color{blue}{\subsection{Contributions}}}

The focus of this article is to answer the question of \emph{how to design a building block that synchronizes intermittently-powered \replace{wireless embedded}{CRFID} systems?} To this end, we investigate the WISP platform and provide initial observations and limitations pertaining to the synchronization of these devices. In particular, the main contributions and findings of 
this article are:
%
\begin{itemize}
	\item To the best of our knowledge, this is the first study that focuses on the synchronization of tags in CRFID systems. Even though there are studies on the synchronization of multiple RFID readers, e.g.~\cite{reader_sync}, we are unaware of any study that provides synchronization of individual tags in the communication range of an RFID reader. 
	\item By addressing the aforementioned challenges, we design and implement \emph{the first synchronization primitive} for the tags, namely \emph{WISP-Sync}, inspired by the Proportional-Integral (PI) controllers~\cite{pi2015,yildirim2017adaptive}. WISP-Sync (i) 
	requires only a few computation steps to run efficiently under limited harvested energy and keeps a few variables to hold the synchronization state to recover from power interruptions with minimum overhead---\emph{addressing Challenge I}; (ii) is adaptive to react to clock instabilities in a fast manner---\emph{addressing Challenge II}.
	%
	\item We provide theoretical analysis to prove that this primitive establishes synchronization and to reveal the factors affecting its synchronization performance.
	%
	\item Evaluations in our testbed show that an average synchronization error of approximately 1\,ms can be ensured among tags with an energy overhead of 1.85\,$\mu$J per synchronization round using this  primitive.
\end{itemize}
%
We note that WISP-side implementations throughout the article are done using C and assembly language. All source codes and scripts used to generate the results in the article (including parsing, post-processing and measurement results) are available via~\href{https://goo.gl/RuWIZ2}{https://goo.gl/RuWIZ2}.

@@

\section{Synchronizing Battery-Free {\color{blue}CRFIDs}}
\label{sec:Related-Work}

@@

\subsection{Synchronization in Conventional WSNs}
\label{sec:WSN-Sync}

@@

\subsubsection{{\color{blue}State-of-the-art Synchronization Techniques}}

The most common synchronization mechanism is to propagate the time information of a particular \emph{reference node} to let receiver nodes synchronize themselves to the received reference time information. Commonly, \emph{least-squares regression} is employed by several practical synchronization protocols to adjust the offset and the frequency of the software clocks ~\cite{Maroti2004,Lenzen:2015,TATS:2016}. On the other hand \emph{iterative} computation methods, e.g. in ~\cite{Yildirim:AVTS:2014,pi2015,yildirim2017adaptive,Yildirim:Gradient:2016}, perform identical computation steps at each re-synchronization round. With iterative methods, sensor nodes establish synchronization after a finite number of rounds in a lightweight manner in terms of computation and memory requirements. There are also \emph{fully distributed} approaches in which sensor nodes interact only with and synchronize to their direct neighbors in a peer-to-peer fashion without the requirement of a dedicated reference node. In these approaches, nodes employ computation methods based on \emph{distributed consensus}~\cite{Sommer2009Gradient,SchenatoFiorentin:2011,Yildirim:External:2014}. \add{Since the prominent factor affecting the clock frequencies in WSNs is the temperature~\cite{schmid2010case}, there are also temperature compensated time synchronization techniques, e.g. \cite{Schmid:2009}. However, such techniques require a calibration phase in order to build a temperature--clock frequency table.}

\subsection{\color{blue}Fundamental Challenges of Synchronizing CRFIDs}
\label{sec:CRFIDs}

We now delve into the synchronization characteristics of \remove{RF-powered wireless embedded systems. To this end, we will consider the} WISP, the de facto CRFID platform. The main aspects pertaining to synchronization in the WISP platform can be stated as follows:

\subsubsection{Single-hop reader-tag architecture} WISP tags are deployed inside the communication range of an RFID reader and they can communicate only with the reader using backscatter communication. Therefore, the RFID reader itself is the natural reference device to establish synchronization among the WISPs, promoting \emph{reader-tag synchronization}. Since WISP tags are unable to communicate with their neighboring nodes directly using backscatter communication as of now, \emph{tag-tag synchronization} is impossible.

\begin{figure}
	\centering
	\includegraphics[width=0.85\columnwidth]{figures/clock_frequency_dc_rf.eps}
	\caption{\label{fig:frequency_dc_rf} In our testbed, see Section~\ref{sec:testbed} for the details, we programmed a single WISP tag so that it toggles one of its output port whenever its 32~KHz built-in timer progresses 32000 clock ticks. In the ideal case where the crystal oscillator of the tag does not drift, the tag would toggle its output port at each 1 second, i.e. with a frequency of 1\,Hz. During our experiment under a stable environmental temperature, we connected the output port of the tag to a logic analyzer and measured the frequency of the toggle event. We plotted our measurements when the tag is powered by the Flash Emulation Tool (FET) with a stable DC power and when the tag is powered by the RFID Reader at a distance of 5~cm. \remove{Since harvested RF energy was different and unstable than the constant DC case, the frequency of the oscillator and in turn the toggle event was quite different. }We measured the mean and standard deviation of the frequency in the DC case as 1.1653 and 0.00028553, and these values for the RF case were 1.1793 and 0.00026049, respectively. }
\end{figure}

\subsubsection{Continuously varying voltage level} In WSNs, the battery level decreases gradually that allows stable voltage levels in the short term. On the contrary, on the WISP platform, the fluctuating input voltage prevents short-term stability of the clock hardware and introduces significant drift. Hence, the prominent factor affecting the frequency of the crystal oscillator is the varying voltage level\remove{ rather than the temperature of the WISP tags}, see Fig.~\ref{fig:frequency_dc_rf}.

\subsubsection{Frequent loss of synchronization state} Contrary to sensor nodes, WISP tags frequently ``die'' due to power loss and they need to save the synchronization state, e.g. the clock offset and relative frequency, into the non-volatile memory to recover when they harvested sufficient energy to start up. However, saving computational state is also an energy consuming task~\cite{blisp}.

\subsubsection{Computation and memory overhead sensitivity} The classical motto of WSNs, ``compute instead of communicate whenever possible''~\cite[p. 44]{karl2007protocols}, is no longer valid for the WISP platform since backscatter communication comes almost for free~\cite{ekhonet}. Due to the intermittent power, lightweight methods in terms of computation and memory are required for the synchronization. Since methods like least-squares regression are computationally heavy and require considerable amount of memory~\cite{pi2015} they should be avoided.

\subsubsection{Limitations of the EPC Gen 2 standard} The WISP firmware implements the EPC Gen 2 standard~\cite{epc_gen2} which increases the compatibility with the existing RFID systems. However, the standard introduces limitations, e.g. currently it does not assign timestamps to the radio packets, which is a fundamental requirement to establish synchronization. Moreover, communication delays between the reader and tag are quite dependent on the implementation of this standard by RFID readers. Unfortunately, these issues lead to less accurate synchronization as compared to existing WSN solutions, as shown by our measurements presented in the following sections.

@@

\subsection{Low Power Operation Modes and The Clock System}
\label{subsec:clock_system}

@@

\begin{figure}
	\centering
	\begin{tikzpicture}
	\node[anchor=south west, inner sep=0] at 
	(0,0){\includegraphics[width=\columnwidth]{figures/testbed_resized.jpg}};

	\node [fill= white,circle] at (7,2) { };
	\node [color=black ] at (7,2) { 1 };

	\node [fill= white,circle] at (8,4) {  };
	\node [color=black ] at (8,4) { 2 };

@@

	\node [fill= white,circle] at (5.1,2) {  };
	\node [color=black ] at (5.1,2) { 4 };

	\node [fill= white,circle] at (6.5,4.5) {  };
	\node [color=black ] at (6.5,4.5) { 5 };

@@

In this section, we provide a mathematical framework that models the built-in clocks and in particular the clock drifts of the WISP tags and the RFID reader. We will use these models to prove the correctness of our synchronization algorithms and to derive mathematical expressions that represent their synchronization performance.
\begin{figure*}
	\centering
	\includegraphics[width=1.4\columnwidth] {figures/relative_clock_frequency.eps}
	\caption{\label{fig:frequency}The frequencies of the toggle events (obtained using the same steps presented in Fig.\ref{fig:frequency_dc_rf}) of two RF-powered WISP tags, placed at 5 cm distance from the RFID reader. On the left the individual event frequencies ($f_1$ and $f_2$) are presented, and their relative clock frequency ($\frac{f_1}{f_2}-1$) is presented on the right. We measured the mean and standard deviation of $f_1$ as 1.1793 and 0.00026049 and those of $f_2$ as 1.1705 and  0.00068239, respectively.}
\end{figure*}
In this framework, we denote a single WISP tag by $w$ and the RFID reader by $r$. The Timer B0 mentioned in Section~\ref{subsec:clock_system} can be considered as the built-in clock, i.e. \emph{local clock}, of the tag $w$, 
which we denote by $C_{w}$. The value of $C_{w}$ at time $t$ can be modeled as
%
\begin{eqnarray}
C_{w}(t) & \triangleq & C_w(t_0)+ \int_{t_0}^{t}f_{w}(u)du, 
\label{eq:local_clock}
\end{eqnarray}
%
where $t_0$ represents the time at which tag $w$ is powered on and  $f_{w}(u)$ represents the \emph{instantaneous oscillator frequency} of $C_{w}$ at time $u$. Since the frequency of the oscillators are affected by environmental factors such as temperature and voltage level, they are not stable~\cite{Lenzen2009Optimal} and $f_{w}$ is not a constant but it is \emph{time varying}, see Fig.~\ref{fig:frequency}. From \eqref{eq:local_clock}, it can be noticed that $C_w$ is \emph{unitless}. Indeed, it can be seen as a counter of oscillation events that are produced at a frequency of $f_w$.

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/Sender-Receiver.pdf}
	\caption{\label{fig:sender_receiver} The message exchange among the host computer, the RFID reader and the tag for sender-receiver synchronization. The host machine, i.e. a PC,  sends the high level commands to the reader via LLRP \emph{AccessSpec} message and receives the results through an \emph{ASReport}. The RFID reader follows the steps defined in the EPC Gen2 standard: (i) performs a \emph{Handshake} to initialize the communication with the active tag; (ii) performs a \emph{Command Initialization} (Cmd Init) by requesting a random number from the tag (Req\_n) and receiving the random number (newRN); (iii)  performs a \emph{Read} command by sending its request and receiving the timer value. The reader assigns the \emph{FirstSeenTimestamp} to the tag at time $t_1$ and the tag timestamps the command reception event at time $t_2$ with its local clock reading.}
\end{figure}

@@

\subsection{Transmission Delay and Jitter Measurements}
